Analyzing Network Transfer Protocols on a Simulated Interplanetary Internet

### 2024-04-10 paper/paper.tex
added data analysis

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. (See
appendix) All entities are able to block signals if the ray passes through the
radius of the entity. One of the largest factors contributing to transmission
error is the distance travelled by the electromagnetic signal between
transmitting and receiving nodes. By generalizing Friis' transmission
formula\cite{Friis} for isotropic emitters and receivers, we can obtain the
ratio of signal intensity received to signal intensity transmitted, known as
Free Space Path Loss (FSPL). We can use FSPL as our transmission error rate,
giving the following expression: $Error Rate = {(\frac{4 \pi d}{\lambda})}^2$
where $d$ is the distance between nodes in metres and $\lambda$ is the
wavelength of the electromagnetic wave used in transmission. As much of space
communications utilizes the Ka-band of frequencies (27--40
GHz)\cite{Morabito_Hastrup} we opted to use a transmission frequency of 30Ghz.
The transmission wavelength can be then calculated according to
$\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

@@

We used NS-3~\cite{ns-3} to implement a simulation of this network. Our final
design consisted of a set of routers representing the different entities in our
physics simulation which were each linked to every other router using a
point-to-point channel. Essentially, this allowed us to have a wired network
with configurable delay and error along with the ability to take certain
channels down if they are obstructed that we could make behave like an
interplanetary network would without having to deal with the complexities of
configuring wireless communication. In addition to the mesh of routers, we added
two individual nodes to the network with zero delay that can be connected to any
one router in the network. One of these two nodes acts as a sender and the other
one acts as a receiver. This configuration of sender and receiver as separate
nodes made it easier to decouple the applications from the network topology and
allowed us to programatically create arbitrary topologies.

@@

Data was scraped from the trace output that was generated while running the 
ns3 simulation. Trace data was given when a message was queued, dequeued, 
and received at the destination. This data also included the amount of 
data being sent in bytes. There is other data on the messages being sent, but 
was not tracked for the purpose of this project. The trace data was parsed 
using regex expressions to extract the needed numbers from the string.
This data was then fed into a pandas dataframe. A unique id had to be 
generated for each message since once a message was recieved the id was 
reused. This was done by tracking when each message was received with id 
$x$ and after incrementing a number by one, say $y$. The unique id was then 
created by hashing $xy$. Once the data frame was created it had to be merged 
on itself to have one row of the dataframe represent exactly one message rather 
then one row be either queue, dequeue, or received.

@@

UDP had an average time from being queued to received of
195.4 seconds, an average time being queued of
18.36 seconds, and an average message time in transit
177.0 seconds. The average amount of data sent was 
1051.6 bytes. Notice the scientific notation in the 
corner of the graph. The graph shows a large amount of 
sensitivity to the motion of the planets which can be 
seen by the curvature of the graph. At time around 28000
seconds, the topology suddenly changed. Caused by a satellite
blocking the path between Mars and Earth. This caused a secondary, 
less efficient path to be taken. This shows why there are several 
"layers" of time which it took. They likely represent the different paths 
which the message took before being sent to its destination. Each quanta in
the time it took to complete the message 
represents the smaller jump it takes to get from a satellite to its 
orbiting entity. The increase in time likely represents that while the 
direct Earth and Mars link was down the smaller satellites were actually orbiting 
away from the message destination.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/tcp.png}
  \includegraphics[width=0.4\textwidth]{media/tcp_data.png}
  \caption{UDP analyzed}
\end{figure}

>>>>

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. (See
appendix) All entities are able to block signals if the ray passes through the
radius of the entity. One of the largest factors contributing to transmission
error is the distance traveled by the electromagnetic signal between
transmitting and receiving nodes. By generalizing Friis' transmission
formula\cite{Friis} for isotropic emitters and receivers, we can obtain the
ratio of signal intensity received to signal intensity transmitted, known as
Free Space Path Loss (FSPL). We can use FSPL as our transmission error rate,
giving the following expression: $Error Rate = {(\frac{4 \pi d}{\lambda})}^2$
where $d$ is the distance between nodes in metres and $\lambda$ is the
wavelength of the electromagnetic wave used in transmission. As much of space
communications utilizes the Ka-band of frequencies (27--40
GHz)\cite{Morabito_Hastrup} we opted to use a transmission frequency of 30Ghz.
The transmission wavelength can be then calculated according to
$\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

@@

We used NS-3~\cite{ns-3} to implement a simulation of this network. Our final
design consisted of a set of routers representing the different entities in our
physics simulation which were each linked to every other router using a
point-to-point channel. Essentially, this allowed us to have a wired network
with configurable delay and error along with the ability to take certain
channels down if they are obstructed that we could make behave like an
interplanetary network would without having to deal with the complexities of
configuring wireless communication. In addition to the mesh of routers, we added
two individual nodes to the network with zero delay that can be connected to any
one router in the network. One of these two nodes acts as a sender and the other
one acts as a receiver. This configuration of sender and receiver as separate
nodes made it easier to decouple the applications from the network topology and
allowed us to programmatically create arbitrary topologies.

@@

Data was scraped from the trace output that was generated while running the 
ns3 simulation. Trace data was given when a message was queued, dequeued, 
and received at the destination. This data also included the amount of 
data being sent in bytes. There is other data on the messages being sent, but 
was not tracked for the purpose of this project. The trace data was parsed 
using regex expressions to extract the needed numbers from the string.
This data was then fed into a pandas dataframe. A unique id had to be 
generated for each message since once a message was received the id was 
reused. This was done by tracking when each message was received with id 
$x$ and after incrementing a number by one, say $y$. The unique id was then 
created by hashing $xy$. Once the data frame was created it had to be merged 
on itself to have one row of the dataframe represent exactly one message rather 
then one row be either queue, dequeue, or received.

@@

% UDP had an average time from being queued to received of
% 195.4 seconds, an average time being queued of
% 18.36 seconds, and an average message time in transit
% 177.0 seconds.

The average amount of data sent was 
1051.6 bytes. The number of bytes 
Notice the scientific notation in the 
corner of the graph.

% remove this line if stats are re-calculated
The average time that a message was in transit was close to 216 seconds.
The time spent waiting in queue was 0.094\% of the time it took for a message to completely 
transmit.

The graph shows a large amount of 
sensitivity to the motion of the planets which can be 
seen by the curvature of the graph. At time around 28000
seconds, the topology suddenly changed. Caused by a satellite
blocking the path between Mars and Earth. This caused a secondary, 
less efficient path to be taken. This shows why there are several 
"layers" of time which it took. They likely represent the different paths 
which the message took before being sent to its destination. Each quanta in
the time it took to complete the message 
represents the smaller jump it takes to get from a satellite to its 
orbiting entity. The increase in time likely represents that while the 
direct Earth and Mars link was down the smaller satellites were actually orbiting 
away from the message destination. UDP messages spent much more time being queued than 
the other messages. This may imply that messages were being transferred back and forth 
between two closely orbiting objects, which could explain the many different layers 
which it can take for a message to be sent. It could also be that since the average 
message is larger than TCP or New Reno it takes longer for the message to be sent 
and it is transmitting.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/tcp.png}
  \includegraphics[width=0.4\textwidth]{media/tcp_data.png}
  \caption{TCP analyzed}
\end{figure}

The average amount of bytes transferred is 394 bytes. This is likely from 
the difference in the amounts of bytes transferred in an ACK message compared 
to when data is being sent.

Unlike in UDP the motion of the planets is not observable in the graph. However this is 
because there are several different paths which the TCP message could have taken which can be seen. 
Since TCP requires a higher amount of base time to establish a connection compared to UDP, this 
means that when it "hops" between entities it adds a larger base time.

At the start of the time we can see 4 separate paths. This would be the different times that it would take 
for any of the 4 paths between Earth and Mars, that is Earth to the ISS to the Mars Orbiter then to Mars,
Earth to the Mars Orbiter to Mars, Earth to the ISS to Mars, and finally Earth to Mars. After one of the 
entities is blocking a direct signal it becomes only that 3 of these options possible, with likely only two actual 
paths being viable. With less available 
bandwidth it becomes slower on average. Smaller time may also be correlated to the minority of messages which were
sent with a smaller message size.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/tcp_future.png}
  \includegraphics[width=0.4\textwidth]{media/tcp_future_data.png}
  \caption{TCP in the future analyzed}
\end{figure}

The second set of graphs show TCP in some time in the future. We can see that the time that messages
take to transfer does in fact change over time as the entities change distances. We can also see a similar
pattern of behaviour changing as some topology changes in the model.

\subsection{New Reno}

New Reno, a modified version of TCP used to help elevate congestion ~\cite{Nahar2016}, was used 
in place of SCP TP. New Reno behaved very similar to TCP. Its average data 
transfer was 407.5 bytes.

It preformed marginally better in its time from being queued to the message received. This 
was mostly due to a decrease in time that it was in queue, which is to be expected. There was 
a very marginal increase in time in transit however.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{media/new_reno.png}
  \includegraphics[width=0.4\textwidth]{media/new_reno_data.png}
  \caption{TCP in the future analyzed}
\end{figure}

Similar to TCP we can see the two stages which occur in the simulation. One where most of 
the messages were likely being routed from directly from Earth to Mars, which then gets transformed 
into having to route through one of the external satellites orbiting each entity, which decreases the 
average time it takes to transmit the message.

\subsection{Protocol Comparison}

Since a large amount of propagation delay dominates the amount of time that it takes for messages total 
transmit time from being queued to being received at the appropriate node, other increases in performance 
mean only a slight decrease in time sending the message. An important statistic to pay attention to is that 
UDP almost always sent a message with a size of 1052 bytes. This means that even though TCP, New Reno, and 
UDP had sent a similar number of messages, the actual throughput for UDP was significantly higher, which 
is much more significant in an IPN situation. In the data graphs we can see that not only was the maximum 
data being sent was less than UDP, but also had to send messages of very little size, most likely acknowledgements,
as well. New Reno was seen to improve on the queuing time compared to both TCP and UDP. Which would mean in a 
high reliability situation and in situations where higher traffic is sent through a network with a relatively low 
bandwidth, like a IPN, where satellites are limited by how much data they can transmit, New Reno's effect can be 
more significant. This should only be used in a connection where propagation delay is less of an issue like 
between planet and satellite communication. Queuing time in UDP took a more significant percentage of its 
time than both TCP and New Reno. In general however, UDP is the best option for primary communication between 
Mars and Earth communication.

\section{Future Work}

Future work that could be done would be to implement or partially implement SCP TP as a network 
protocol which could be used and compared to the protocols analyzed in this paper. Other protocols 
like the Linklider or load and store protocols could also be implemented, to create a more realistic 
IPN.

The physics engine could be expanded to included 3D orbits, particularly around planets, where networks of 
satellites could be created to create a proper level of communication that is maintained today, with geocentric 
and polar orbiting satellites. Connection points from a planet to off the planet could have a direction which it 
needs to send a message in for it to be able to connect. For example a rover on Mars in our simulation could technically 
send a message in any direction, even if they were sending it through the surface of the planet.The error model in 
the physics engine could be expanded to include interference like solar radiation, atmosphere, and other interference.

Other stats could be extracted from the NS-3 model such as packet loss, message path, handshake time. It could expand 
so that the network was not just focused on Earth to Mars communication. It could be expanded to include other celestial 
objects in the solar system, as well as potential future objects.

The NS-3 simulation could dynamically choose the network protocol to use based on the level of error, message size,
estimated propagation error, and other factors. This could build a model which uses not only one protocol but all of them 
and leverages the advantages of each when it is necessary. This combined with a more sophisticated model of the solar 
system would create a very competent simulation of the current operation of the IPN, and could then be expanded or tweaked 
to find if expansions to it could be handled properly. This would need a combination of all of the suggestions above 
to be successful.

\section{Conclusion}

In this paper we outlined how we created a model of an interplanetary internet (IPN) using a physics engine which fed into 
a NS-3 simulation. The physics simulation used a simplified model of orbiting planets and satellites as well as physics calculations 
to create data which would be the output for any input time $t$. The NS-3 model used a point to point base which programmatically
took the data from the physics engine and created nodes which acted as routers. A simulation was ran which transfered data between 
Earth and Mars, using satellites and other entities to relay messages around blocked connections. Analysis was done on three
different network protocols used in the IPN, that is UDP, TCP and New Reno. 
UDP was found to have the most throughput out of all of the three protocols, while New Reno reduced the amount of time a message 
was queued.
